An Executable Formal Semantics For A Functional Actor Language

### 2017-12-20 paper/paper.tex
added all citations

\begin{abstract}
A simple reference actor language has been proposed in~\cite{}, serving as a
sound mathematical calculus foundation for various actor languages used in
distributed applications that involves concurrency. Even though the authors
of~\cite{} provided in the paper an operational semantics of the actor language,
an executable implementation was not given. In this project, we will give the
reference actor language an implementation that by construction conforms to its
formal operational semantics using the K framework~\cite{}, a rule-based
semantic framework. We will show that our executable formal semantics are
capable of executing some meaningful actor systems examples. We hope our
executable formal semantics can be served as a starting point for future work in
developing automatic equivalence provers using the K framework.
\end{abstract}
%
\section{Introduction}
The K framework~\cite{} is a framework for defining executable operational
semantics of a programming language. K is based on rewriting logic, organizing
the state of a program into cells and rewriting the particular parts of the
configuration. K also has access the entire computation at any particular
time. These features and more allow for difficult language constructs to be
designed with ease in K. On top of that, K provides backend tooling for
additional features based on the semantics of a programming language, such as a
semantic debugger and a full formal verifier. See Figure~\ref{fig:k-tools} for a
complete view of the tools one gets from K based off an operational semantics.

@@

\begin{figure}
          \centering
          \includegraphics[scale=0.13]{../k-overview.png}
	  \caption{Backend tools K provides from just an operational
            semantics. Some tools such as a semantics-based compiler and
            test-case generation are in progress.}
          \label{fig:k-tools}
\end{figure}
To demonstrate the power of K, languages as complex as C, Java, and JavaScript
have been fully defined using the K framework~\cite{c,java,jscript}. On top of that, in
all of these languages and more, nontrivial programs have been formally verified
using K's prover, such as proofs involving AVL trees and red-black trees. We
would like to add an actor-based language to the growing list of languages
defined in K, so we can take advantage of K's formal verification engine and
more.

Semantics of actor-based languages have been defined on paper, for example
in~\cite{}. Additionally, there are many implementations of actor-based systems,
including Erlang~\cite{}, Akka~\cite{}, Scheme~\cite{}, and much more. However,
there is currently no formally defined \emph{executable} operational semantics
of an actor-based language. In K, there was previous work done to define the
semantics of Orc, a concurrent programming language~\cite{}, though that work is
incomplete.

In this paper, we aim to define the semantics of the actor language proposed
in~\cite{}.  We choose this language as the semantics are clearly defined in the
paper, and it is one of the more simple actor frameworks. Defining this language
in K will give assurance to the correctness of the language - as K is executable
this can be tested much easier than defining a language on paper. It will also
provide more assurance on K's language-defining ability, as well as open up to
formal analysis using the tools seen in Figure~\ref{fig:k-tools}.

@@

For the branching, we would like to only specify that only the first argument
should reduce to a value, to avoid reduction in the branch we end up not taking.
In the actor semantics from~\cite{}, this is accomplished with the following:
\[ \texttt{if}(e_0, e_1, e_2) \;\;\;\text{abbreviates}\;\;\; \texttt{app}(\texttt{br}(e_0, \lambda z.e_1, \lambda z.e_2), \texttt{nil}) \;\;\;\text{for $z$ fresh} \]
Notice that the second and third arguments in the branch construct are lambda
expressions, which are values, and thus will not be reduced any farther. In our
K semantics, we emulate this, but we could also define \texttt{br} directly to
be only strict in the first argument as follows:
\begin{verbatim}
syntax Exp ::= "br" "(" Exp "," Exp "," Exp ")" [strict(1)]
\end{verbatim}
Strictness constructs like this avoid unnecessary syntactic sugar, and cleanly
specify most reduction contexts one would require.

@@

In this paper we showed how to define the actor semantics from~\cite{} using the
K framework. We described in detail the syntax and semantics of the language in
K, as well as the configuration for our actor language. We also showed how we
can use K's strictness attributes to emulate the reduction contexts described
in~\cite{}. Finally, we discussed a complicated actor program, the tree product
example, and its encoding in K.

One immediate area for future work includes defining more actor programs and
more in-depth actor programs in K. Having a large test suite of complicated
examples can be used to ensure that the semantics in K and the semantics
from~\cite{} do not have any bugs. It can also be used to determine rule
coverage, affirming that all semantic rules can be applied.

Another important area for future work involves taking advantage of K's
backend tools for formal analysis, such as proving actor programs correct using
K's verification engine. For verification, K uses reachability logic~\cite{}, a
language-independent formal verification framework. Reachability logic supports
all-path reachability, so it can be used to verify programs involving concurrent
semantics. However, in order to fully take advantage of this verification, one
would need to implement partial-order reduction in K in order to reduce the
state space.

>>>>

\begin{abstract}
A simple reference actor language has been proposed in~\cite{actor}, serving as a
sound mathematical calculus foundation for various actor languages used in
distributed applications that involves concurrency. Even though the authors
of~\cite{actor} provided in the paper an operational semantics of the actor language,
an executable implementation was not given. In this project, we will give the
reference actor language an implementation that by construction conforms to its
formal operational semantics using the K framework~\cite{kframework}, a rule-based
semantic framework. We will show that our executable formal semantics are
capable of executing some meaningful actor systems examples. We hope our
executable formal semantics can be served as a starting point for future work in
developing automatic equivalence provers using the K framework.
\end{abstract}
%
\section{Introduction}
The K framework~\cite{kframework} is a framework for defining executable operational
semantics of a programming language. K is based on rewriting logic, organizing
the state of a program into cells and rewriting the particular parts of the
configuration. K also has access the entire computation at any particular
time. These features and more allow for difficult language constructs to be
designed with ease in K. On top of that, K provides backend tooling for
additional features based on the semantics of a programming language, such as a
semantic debugger and a full formal verifier. See Figure~\ref{fig:k-tools} for a
complete view of the tools one gets from K based off an operational semantics.

@@

\begin{figure}
          \centering
          \includegraphics[scale=0.13]{../k-overview.png}
	  \caption{Backend tools K provides from just an operational
            semantics. Some tools such as a semantics-based compiler and
            test-case generation are in progress.}
          \label{fig:k-tools}
\end{figure}
To demonstrate the power of K, languages as complex as C, Java, and JavaScript
have been fully defined using the K framework~\cite{kc,kjava,kjs}. On top of that, in
all of these languages and more, nontrivial programs have been formally verified
using K's prover, such as proofs involving AVL trees and red-black trees. We
would like to add an actor-based language to the growing list of languages
defined in K, so we can take advantage of K's formal verification engine and
more.

Semantics of actor-based languages have been defined on paper, for example
in~\cite{actor}. Additionally, there are many implementations of actor-based systems,
including Erlang~\cite{erlang}, Akka~\cite{akka}, Scheme~\cite{scheme}, and much more. However,
there is currently no formally defined \emph{executable} operational semantics
of an actor-based language. In K, there was previous work done to define the
semantics of Orc, a concurrent programming language~\cite{orc}, though that work is
incomplete.

In this paper, we aim to define the semantics of the actor language proposed
in~\cite{actor}.  We choose this language as the semantics are clearly defined in the
paper, and it is one of the more simple actor frameworks. Defining this language
in K will give assurance to the correctness of the language - as K is executable
this can be tested much easier than defining a language on paper. It will also
provide more assurance on K's language-defining ability, as well as open up to
formal analysis using the tools seen in Figure~\ref{fig:k-tools}.

@@

For the branching, we would like to only specify that only the first argument
should reduce to a value, to avoid reduction in the branch we end up not taking.
In the actor semantics from~\cite{actor}, this is accomplished with the following:
\[ \texttt{if}(e_0, e_1, e_2) \;\;\;\text{abbreviates}\;\;\; \texttt{app}(\texttt{br}(e_0, \lambda z.e_1, \lambda z.e_2), \texttt{nil}) \;\;\;\text{for $z$ fresh} \]
Notice that the second and third arguments in the branch construct are lambda
expressions, which are values, and thus will not be reduced any farther. In our
K semantics, we emulate this, but we could also define \texttt{br} directly to
be only strict in the first argument as follows:
\begin{verbatim}
  syntax Exp ::= "br" "(" Exp "," Exp "," Exp ")" [strict(1)]
\end{verbatim}
Strictness constructs like this avoid unnecessary syntactic sugar, and cleanly
specify most reduction contexts one would require.

@@

In this paper we showed how to define the actor semantics from~\cite{actor} using the
K framework. We described in detail the syntax and semantics of the language in
K, as well as the configuration for our actor language. We also showed how we
can use K's strictness attributes to emulate the reduction contexts described
in~\cite{actor}. Finally, we discussed a complicated actor program, the tree product
example, and its encoding in K.

One immediate area for future work includes defining more actor programs and
more in-depth actor programs in K. Having a large test suite of complicated
examples can be used to ensure that the semantics in K and the semantics
from~\cite{actor} do not have any bugs. It can also be used to determine rule
coverage, affirming that all semantic rules can be applied.

Another important area for future work involves taking advantage of K's
backend tools for formal analysis, such as proving actor programs correct using
K's verification engine. For verification, K uses reachability logic~\cite{reachabilitylogic}, a
language-independent formal verification framework. Reachability logic supports
all-path reachability, so it can be used to verify programs involving concurrent
semantics. However, in order to fully take advantage of this verification, one
would need to implement partial-order reduction in K in order to reduce the
state space.
